A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2017-02-02 rigid_bodies.tex
nvt review
canonical review

In this context, the Hamiltonian of the rigid body can be rewritten as
\begin{equation}
\label{eq:H_split_omega}
\mathcal{H} =  U(\vt r, \vt q) +  \sum_{i=1}^N \left( \frac{\tr{\vt p_i} \vt p_i}{2m_i} + \frac{1}{2} \sum_{k=1}^3 I_{ik} \omega_{ik}^2\right),
\end{equation}
from which we recognize partial Hamiltonians, each of one generates equations of motion with known analytical solutions. In particular, we have $\mathcal{H}_b = U(\vt r, \vt q)$, $\mathcal{H}_{t} = \frac{1}{2m}\tr{\vt p} \vt p$, and $\mathcal{H}_k = (1/2) I_k \omega_k^2$. Of coarse, $\mathcal{H} = \mathcal{H}_b + \mathcal{H}_t + \sum_{k=1}^3 \mathcal{H}_k$. The NVE dynamics of this system is governed by a Liouville operator given by\cite{Abreu_2017}
\begin{equation}
\label{eq:full_operator}
i\!L_{NVE} = i\!L_t + i\!L_b + i\!L_r,
\end{equation}
where the individual operators are defined as follows:
\[
\begin{split}
&boost:\\
i\!L_b &= \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right), \\
&translation: \\
 i\!L_t &= \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i}, \quad \text{and} \\
&uniaxial \; rotation:\\
 i\!L_k &= \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right),
\end{split}
\]
using $i\!L_r = \sum_{k=1}^3 i\!L_k$ for a complete torque-free rotation.
The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
\begin{equation}
\label{eq:trotter_splitting_NVE}
e^{\Delta t i\!L_{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b},
\end{equation}
Update of forces and torques is required only after the free translation and free rotation have been carried out. Following Miller \textit{et al}.,\cite{Miller2002} each complete rotation is splitted as
\begin{equation}
\label{eq:splitting_rot}
e^{\Delta t i\!L_r} =  e^{\frac{\Delta t}{2} i\!L_3} e^{\frac{\Delta t}{2} i\!L_2} e^{\Delta t i\!L_1} e^{\frac{\Delta t}{2} i\!L_2} e^{\frac{\Delta t}{2} i\!L_3},
\end{equation}
where ${\Delta t}$ is the time step size.

As demonstrated in our previous paper,\cite{Abreu2017}, the action of the propagators 
\begin{align*}
&{\vt r} = {\vt r}_0 + (t/m) {\vt p}_0,
\end{align*}

\begin{align*}
&{\vt p} = {\vt p}_0 + {\vt F}_0 t, \\
&{\vt \pi} = {\vt \pi}_0 +  2 {\mt C}_0 {\vt \tau}_0 t.
\end{align*}

\begin{align*}
&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \ \text{and} \\
&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0, 
\end{align*}

@@

\subsection{Reversible, Measure-Preserving Integration of the Canonical Equations of Motion}

As in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999, Tuckerman2001} Thus, we can employ the Trotter-Suzuki splitting to devise a time-reversible algorithm for integrating the equations of motion. In the microcanonical case, each separate term of the Liouville operator is Hamiltonian and, therefore, each factor of the splitting is a symplectic propagator. As a result, the whole propagator is symplectic and thus preserves the phase-space volume element. It is widely recognized that this property is important for numerical stability.\cite{Skeel1997} In the case of non-Hamiltonian systems, Ezra\cite{Ezra2006} introduced a systematic method for deriving accurate integrators which are not only time-reversible, but also preserve the phase-space measure associated with the original equations of motion. In short, a factorization scheme will preserve a given measure $e^{-w(\vt x)}d\vt x$ if each individual propagator preserves the same measure. As the author has shown,\cite{Ezra2006} for a particular splitting $i\!L = \sum_k i\!L^{(k)}$, measure preservation will take place if $\sum_j \tfrac{\partial}{\partial x_j} [e^{-w(\vt x)}\dot{x}_j^{(k)}] = 0$ for all $k$. Here, $\dot{\vt x}^{(k)}$ refers to the system of equations obtained from the individual operator $i\!L^{(k)}$, which might display its own compressibility $\kappa^{(k)} = \sum_j \partial \dot{x}_j^{(k)}/\partial x_j$. By means of the product rule, it is straightforward to devise an equivalent but more meaningful criterion, which is
\begin{equation}
\label{eq:split_kappa_w}
\kappa^{(k)} = i\!L^{(k)} w.
\end{equation}

@@

 Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$. Note that the term that premultiplies the extended phase-space volume in the invariant-measure above is a function of $\vt \eta$. Given that the time evolution of the physical variables and the thermostat momenta does not depend on $\vt \eta$, we might conclude that, in this case, the measure-preserving condition for every individual operator can actually be relaxed.

\begin{subequations}
	\label{eq:virial_rigid_bodies}
	\begin{align}
	W &= \frac{1}{3} \sum_{i=1}^{N-1} \sum_{j=i+1}^N \tr{\hat{\vt r}}_{ij} \vt F_{ij} \\
	  &= W_\text{a} - \frac{1}{3} \sum_{i=1}^N \sum_{k=1}^{{n_p}_i} {\tr{\vt \delta}_k}_i {{\vt f}_k}_i,
	\end{align}
\end{subequations}
where $W_\text{a}$ is the atomic virial given by Eq.~\eqref{eq:virial_atoms}, while $\vt F_{ij}$ is the total action of body $j$ (and its images) over body $i$ and $\vt f_{ik}$ is the resultant force exerted on the $k$-th atom of body $i$, respectively given by
\
\[
\vt F_{ij} = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \vt f_{ijkl} \quad \text{and} \quad \vt f_{ik} = \sum_{\substack{j=1\\j \neq i}}^N \sum_{l=1}^{{n_p}_j} \vt f_{ijkl}.
\]

\section{Numerical Results}
\label{sec:numerical_results}

In this section, we analyze some performance aspects of the integration schemes when applied to a molecular system in the liquid state. For this, we simulated a system containing 356 TIP3P water molecules\cite{Price2004} at a density of 999.56 $kg/m^3$. The intermolecular interactions were truncated at 9 \AA. The shifted-force (SF) method\cite{Allen1989, Fennell2006, Toxvaerd_2011} was used to treat both the Coulombic and the 12-6 Lennard-Jones (LJ) interactions.

\subsection{Numerical Stability}
\label{sec:performance}

In order to evaluate the stability of the numerical integrators, we employ a measure given by\cite{Tuckerman2010}
\begin{equation}
\label{eq:performance}
D\!E =  \frac{1}{\mathcal N} \sum_{t=1}^{\mathcal N} \left| \frac{E_t - E_0}{E_0} \right|,
\end{equation}
from which it is possible to quantify the average deviation of the supposedly conserved quantity $E$ (either $\mathcal{H}$ of Eq.~\eqref{eq:H_NVE} or $H$ of Eq.~\eqref{eq:H_nvt}) from its initial value $E_0$ after a certain number of time steps $\mathcal N$.
It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to show that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble. \textbf{In order to use a constant value for the smaller time step in the NHC and NHC* operators $(\Delta t/4)$, we factorized those operators analogously to the scheme given by Eq.~\eqref{eq:splitting_rot}}. Recalling that in the NVT dynamics the conserved quantity is the extended energy $H$, in Table~\eqref{table:denvt} we show the results of Eq.~\eqref{eq:performance} as well as the average temperatures obtained in each case. It can be seen that both integrators, represented by the Eq.~\eqref{eq:trotter_splitting_NHC} and Eq.~\eqref{eq:modified_splitting}, systematically achieve the specified temperature, regardless of the time step size. In addition, the energy drift is, at its worst, around $1 \%$.

\subsection{New Results}

@@

\subsection{Partition of Energy}
\label{sec:energypartition}

\subsection{Sampling Validation}
\label{sec:samplingvalidation}

@@

>>>>

In this context, the Hamiltonian of the rigid body can be rewritten as
\begin{equation}
\label{eq:H_split_omega}
\mathcal{H} =  U(\vt r, \vt q) +  \sum_{i=1}^N \left( \frac{\tr{\vt p_i} \vt p_i}{2m_i} + \frac{1}{2} \sum_{k=1}^3 I_{ik} \omega_{ik}^2\right),
\end{equation}
from which we recognize partial Hamiltonians, each of one generates equations of motion with known analytical solutions. In particular, we have $\mathcal{H}_b = U(\vt r, \vt q)$, $\mathcal{H}_{t} = \frac{1}{2m}\tr{\vt p} \vt p$, and $\mathcal{H}_k = (1/2) I_k \omega_k^2$. Of coarse, $\mathcal{H} = \mathcal{H}_b + \mathcal{H}_t + \sum_{k=1}^3 \mathcal{H}_k$. The NVE dynamics of this system is governed by a Liouville operator given by\cite{Abreu_2017}
\begin{equation}
\label{eq:full_operator}
i\!L_{NVE} = i\!L_t + i\!L_b + i\!L_r,
\end{equation}
where the individual operators are defined as follows:
\[
\begin{split}
&boost:\\
&i\!L_b = \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right), \\
&translation: \\
& i\!L_t = \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i}, \quad \text{and} \\
&uniaxial \; rotation:\\
& i\!L_k = \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right),
\end{split}
\]
using $i\!L_r = \sum_{k=1}^3 i\!L_k$ for a complete torque-free rotation. As explained in our previous paper,\cite{Abreu2017} those individual operators can be applied analytically according to
\[
\begin{split}
&boost:\\
&{\vt p} = {\vt p}_0 + {\vt F}_0 t \\
&{\vt \pi} = {\vt \pi}_0 +  2 {\mt C}_0 {\vt \tau}_0 t\\
&translation:\\
&{\vt r} = {\vt r}_0 + (t/m) {\vt p}_0\\
&uniaxial \; rotation:\\
&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0
\end{split}
\]

Applying the Trotter-susz The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
\begin{equation}
\label{eq:trotter_splitting_NVE}
e^{\Delta t i\!L_{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b},
\end{equation}
Update of forces and torques is required only after the free translation and free rotation have been carried out. Following Miller \textit{et al}.,\cite{Miller2002} each complete rotation is splitted as
\begin{align*}
\label{eq:splitting_rot}
e^{\Delta t i\!L_r} =  e^{\frac{\Delta t}{2} i\!L_3} e^{\frac{\Delta t}{2} i\!L_2} e^{\Delta t i\!L_1} e^{\frac{\Delta t}{2} i\!L_2} e^{\frac{\Delta t}{2} i\!L_3},
\end{align*}
where ${\Delta t}$ is the time step size.

equipartion of energy

@@

\subsection{Reversible, Measure-Preserving Integration of the Canonical Equations of Motion}
\label{sec:numerical_solvers}
As in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999, Tuckerman2001} Thus, we can employ the Trotter-Suzuki splitting to devise a time-reversible algorithm for integrating the equations of motion. In the microcanonical case, each separate term of the Liouville operator is Hamiltonian and, therefore, each factor of the splitting is a symplectic propagator. As a result, the whole propagator is symplectic and thus preserves the phase-space volume element. It is widely recognized that this property is important for numerical stability.\cite{Skeel1997} In the case of non-Hamiltonian systems, Ezra\cite{Ezra2006} introduced a systematic method for deriving accurate integrators which are not only time-reversible, but also preserve the phase-space measure associated with the original equations of motion. In short, a factorization scheme will preserve a given measure $e^{-w(\vt x)}d\vt x$ if each individual propagator preserves the same measure. As the author has shown,\cite{Ezra2006} for a particular splitting $i\!L = \sum_k i\!L^{(k)}$, measure preservation will take place if $\sum_j \tfrac{\partial}{\partial x_j} [e^{-w(\vt x)}\dot{x}_j^{(k)}] = 0$ for all $k$. Here, $\dot{\vt x}^{(k)}$ refers to the system of equations obtained from the individual operator $i\!L^{(k)}$, which might display its own compressibility $\kappa^{(k)} = \sum_j \partial \dot{x}_j^{(k)}/\partial x_j$. By means of the product rule, it is straightforward to devise an equivalent but more meaningful criterion, which is
\begin{equation}
\label{eq:split_kappa_w}
\kappa^{(k)} = i\!L^{(k)} w.
\end{equation}

@@

We end this section by noting that the term that premultiplies the extended phase-space volume in the invariant-measure above is a function of $\vt \eta$. Given that the time evolution of the physical variables and the thermostat momenta does not depend on $\vt \eta$, we might conclude that, in this case, the measure-preserving condition for every individual operator can actually be relaxed. However...the energy drift...

\section{Results}
\label{sec:numerical_results}
 Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$. 
In this section, we analyze the integration schemes introduced in Subsection~\ref{sec:numerical_solvers} and compare the results with those obtained employing the method of Kamberaj~\cite{Kamberaj2005}. For that, we simulated a system containing 356 TIP3P water molecules\cite{Price2004} at a density of 999.56 $kg/m^3$. The intermolecular interactions were truncated at 9 \AA. The shifted-force (SF) method\cite{Allen1989, Fennell2006, Toxvaerd_2011} was used to treat both the Coulombic and the 12-6 Lennard-Jones (LJ) interactions.

It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to show that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble. \textbf{In order to use a constant value for the smaller time step in the NHC and NHC* operators $(\Delta t/4)$, we factorized those operators analogously to the scheme given by Eq.~\eqref{eq:splitting_rot}}. Recalling that in the NVT dynamics the conserved quantity is the extended energy $H$, in Table~\eqref{table:denvt} we show the results of Eq.~\eqref{eq:performance} as well as the average temperatures obtained in each case. It can be seen that both integrators, represented by the Eq.~\eqref{eq:trotter_splitting_NHC} and Eq.~\eqref{eq:modified_splitting}, systematically achieve the specified temperature, regardless of the time step size. In addition, the energy drift is, at its worst, around $1 \%$.

@@

\section{Sampling Validation}
\label{sec:samplingvalidation}

@@

%In order to evaluate the stability of the numerical integrators, we employ a measure given by%\cite{Tuckerman2010}
%\begin{equation}
%\label{eq:performance}
%D\!E =  \frac{1}{\mathcal N} \sum_{t=1}^{\mathcal N} \left| \frac{E_t - E_0}{E_0} \right|,
%\end{equation}
%from which it is possible to quantify the average deviation of the supposedly conserved quantity $E$ %(either $\mathcal{H}$ of Eq.~\eqref{eq:H_NVE} or $H$ of Eq.~\eqref{eq:H_nvt}) from its initial value $E_0$ %after a certain number of time steps $\mathcal N$.
